---
title: "Introduction to aRangodb package"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

In order to understand the examples in the next sections it is desirable understand some basic concepts of the behaving and logic of ArangoDB. Each instance of an ArangoDB server can store multiple databases: even if you don't create any database, there is the default one named **\_system**. Every operation you invoke that is invoked with no database specified affects the **\_system** one. It is suggested to not use this database for applications.

#### References

* **Quick introduction** https://docs.arangodb.com/3.4/Manual/GettingStarted/DatabasesCollectionsDocuments.html
* **Core Concepts** https://docs.arangodb.com/3.4/Manual/DataModeling/
* **AQL** https://docs.arangodb.com/3.4/AQL/

## Install the package <a name="installation"></a>
To install the aRangodb package you can run the following commands.

```{r install, echo=TRUE, include = FALSE, eval=FALSE}
devtools::install_gitlab("krpack/arango-driver")
library(aRangodb)
```

## Usage examples <a name="usage"></a>
In the following sections are shown usage's examples of this driver. The conmplete example, as R script, is located into the repository at the path "examples/arango_full_example.R".
Once installed you have to load the library:

```{r library, echo=TRUE, message=FALSE, warning=FALSE}
library(aRangodb)
```

### Connection to ArangoDB server <a name="connectarango"></a>
 
Connect to an ArangoDB server up and running

```{r connetion_ex, echo=TRUE, eval=FALSE}
arangoConnection <- arango_connection(<instance_ip>, <instance_port>)
```

```{r connection_real, echo=FALSE}
arangoConnection <- arango_connection("138.201.225.72", 41000)

# If you want to delete the existing database to have a clear environment
sandboxArangoDb <- arangoConnection %>% arango_database("sandbox")
if(!is.null(sandboxArangoDb)){
  sandboxArangoDb %>% drop()
}
```


### Create or connect to a database <a name="connectdatabase"></a>

Create a sandbox database: to force creation of a database, if it is not found, use the 
_createOnFail_ option (default is FALSE):

```{r connection_db, echo=TRUE, message=FALSE, warning=FALSE}
sandboxArangoDb <- 
  arangoConnection %>% 
  arango_database("sandbox", createOnFail = TRUE)
```

To use the default database (**DISCOURAGED**) you can do:

```{r connection_default_db, echo=TRUE, message=FALSE, warning=FALSE}
defaultArangoDb <- 
  arangoConnection %>% 
  arango_database()
```

The object that those methods return is an instance of ArangoDatabase class and it will be used for all the operations affecting the database.

### Create or load a collection

From now on all the collections, documents and graphs will be manipulated within "sandbox"
(or the db you had choosen): this means that collections, documents, and graphs will be visible
and available from this DB.
Now create the "person" and "city" collections:

```{r collections, echo=TRUE, message=FALSE, warning=FALSE}
persons <- sandboxArangoDb %>% arango_collection("person", createOnFail = TRUE)
cities <- sandboxArangoDb %>% arango_collection("city", createOnFail = TRUE)
```

Again it has been used the option _createOnFail_ to create the collections if they were not found into the database. The call otherwise would launch an error (collection not found).

### Insertion of documents into a collection

Now we have two different collections: we can populate with some fake data for subsequent examples using the aRangodb::document_insert(key) to create a new document with the given key as identifier, and setting its values using the aRangodb::set() function. The insertion, and the values' updates, are effective once invoked the aRangodb::collection_update() function. It could be a limitation, but for synchronous (and immediate) call there will be a dedicated release of the driver.

```{r collections_insert, echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE}
persons %>% 
  document_insert("john.doe") %>% 
  document_set(age=30, birthday="17/01/1989") %>% 
  collection_update()

persons %>% 
  document_insert("alice.foo") %>% 
  document_set(age=36, birthday="07/02/1983", graduated = TRUE) %>% 
  collection_update()

persons %>% 
  document_insert("brandon.fee") %>% 
  document_set(age=36, birthday="03/11/1983", jobTitle="Software Developer") %>% 
  collection_update()

persons %>% 
  document_insert("charlie.foo") %>% 
  document_set(age=34, birthday="03/02/1985") %>% 
  collection_update()

cities %>% 
  document_insert("London") %>% 
  document_set(position=list(latitude=51.5098, longitude=-2.0122), capital=TRUE, country="UK") %>% 
  collection_update()

cities %>% 
  document_insert("Manchester") %>% 
  document_set(position=list(latitude=53.4839, longitude=-2.2446), capital=FALSE, country="UK") %>% 
  collection_update()
```

```{r collections_insert_true, echo=FALSE, results='hide', eval=TRUE, message=FALSE, warning=FALSE}
persons %>% 
  document_insert("john.doe") %>% 
  document_set(age=30, birthday="17/01/1989") %>% 
  collection_update()

persons %>% 
  document_insert("alice.foo") %>% 
  document_set(age=36, birthday="07/02/1983", graduated = TRUE) %>% 
  collection_update()

persons %>% 
  document_insert("brandon.fee") %>% 
  document_set(age=36, birthday="03/11/1983", jobTitle="Software Developer") %>% 
  collection_update()

persons %>% 
  document_insert("charlie.foo") %>% 
  document_set(age=34, birthday="03/02/1985") %>% 
  collection_update()

cities %>% 
  document_insert("London") %>% 
  document_set(position=list(latitude=51.5098, longitude=-2.0122), capital=TRUE, country="UK") %>% 
  collection_update()

cities %>% 
  document_insert("Manchester") %>% 
  document_set(position=list(latitude=53.4839, longitude=-2.2446), capital=FALSE, country="UK") %>% 
  collection_update()
```

You can use the collection to access some useful information about the collection itself, for example using the getCount() method.

```{r collections_count, echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE}
print(paste("Persons registered:", persons$getCount(), sep = " "))
```

```{r collections_count_real, echo=FALSE, message=FALSE, warning=FALSE}
print(paste("Persons registered:", persons$getCount(), sep = " "))
```

### Working with documents <a name="workingdocuments"></a>

You can use collections to retrieve and access documents within the collection itself. For example you can get ALL the document of a collection (**BE CAREFUL**, collections could contains hundred of results) and access specific document using its key. The aRango::all_documents() return a list with all the documents found for the given collection.

```{r step_1, echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE}
all.cities <- cities %>% all_documents()
all.persons <- persons %>% all_documents()

if(all.cities$London$getValues()$capital){
  print("London is still the capital of UK")
} else {
  print("What's happening there???")
}
```

```{r step_1_real, echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE}
all.cities <- cities %>% all_documents()
all.persons <- persons %>% all_documents()

if(all.cities$London$getValues()$capital){
  print("London is still the capital of UK")
} else {
  print("What's happening there???")
}
```

Each document is an object belonging to the ArangoDocument class: every instance expose the following methods:

* getKeys(), returns a vector of properties for which the document has a valid value (not null)
* getValues(), returns the list of pairs key-value for which exist a not-null mapping


Using a collection object you can filter out documents that match some conditions. To express values
greater or less than some condition to be matched use operators %lt%, %gt%, %leq%, %geq%. Next lines will be translated as _"ehy, give me all cities of UK over latitude 52.0"_

```{r step_2, echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE}
filtered.cities <- cities %>% collection_filter(country="UK", position.latitude %gt% 52.0)

if(is.null(filtered.cities$Manchester)){
  print("Ehy, who moved away Manchester??") # Could be very bad
} else {
  print("Manchester is where should be located")
}

if(!is.null(filtered.cities$London)){
  print("Ehy, who moved away London??") # Could be very bad
} else {
  print("London is where should be located")
}
```

```{r step_2_real, echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE}
filtered.cities <- cities %>% collection_filter(country="UK", position.latitude %gt% 52.0)

if(is.null(filtered.cities$Manchester)){
  print("Ehy, who moved away Manchester??") # Could be very bad
} else {
  print("Manchester is where should be located")
}

if(!is.null(filtered.cities$London)){
  print("Ehy, who moved away London??") # Could be very bad
} else {
  print("London is where should be located")
}
```

You can use the aRangodb::document_set() or aRangodb::document_unset() methods to insert/update or remove an attribute from a document. For example, the following code shown an insertion, followed by an update and an unset.

```{r step_3, echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE}
Lyon <- cities %>% 
  document_insert("Lyon") %>% 
  document_set(capital=FALSE, country="Fran") %>%    # wrong insertion
  collection_update()
  
Lyon %>%
  document_set(country="France") %>%                 # update
  collection_update()
  
Lyon %>%
  document_unset(capital) %>%                        # no more important
  collection_update()
```

### Custom AQL queries <a name="aqlcustom"></a>
If you are interested in custom queries you can use the native Arango Query Language (AQL). To do that
you need to specify the string containing the query using the \@variable_name to indicate some binding
variables: the query will be parsed to check the syntax and then converted to an R function where binding variables are the formal parameters of the function.
Do you want to get all the persons with age more than come threshold?

```{r step_4, echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE}
searchByAgeGreaterThen <- 
  sandboxArangoDb %>% 
  aql("FOR p IN person FILTER p.age > @age RETURN p")
```

```{r step_4_aql, echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE}
searchByAgeGreaterThen <- 
  sandboxArangoDb %>% 
  aql("FOR p IN person FILTER p.age > @age RETURN p")
```

The previous commands affect the current environment by adding a new function, named _searchByAgeGreaterThan_, where formal parameters are the ones corresponding to the variables left unbind in the AQL query, in this case _age_. Before returning a function, the aRango::aql() command execute the parsing of the query, so that syntax error will be highlighted as error. 
In conclusion from now on you can now execute the query as normal R function:

```{r aql_exec, echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE}
filtered.persons <- searchByAgeGreaterThen(age = 30)
```

```{r aql_exec_real, echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE}
filtered.persons <- searchByAgeGreaterThen(age = 30)
```

The results of those functions calls are returned as list, where each result has an automatically generated id within the list (up to now doc<n>, but it can change before the first stable release)

```{r step_5, echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE}
if(length(filtered.persons) != 3){
  print("Did you change something?")
}

print(paste0("'I'm Brandon Fee, isn't it?' ", filtered.persons$doc0$`_key` == "brandon.fee"))
```

```{r step_5_real, echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE}
if(length(filtered.persons) != 3){
  print("Did you change something?")
}

print(paste0("'I'm Brandon Fee, isn't it?' ", filtered.persons$doc0$`_key` == "brandon.fee"))
```

### Working with graphs <a name="workinggrapphs"></a>
Last but not least, you can define a graph structure in the same way you can define a collection.

```{r step_6, echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE}
residenceGraph <- 
  sandboxArangoDb %>% 
  arango_graph("residence", createOnFail = TRUE)
```

```{r step_6_real, echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE}
residenceGraph <- 
  sandboxArangoDb %>% 
  arango_graph("residence", createOnFail = TRUE)
```

If you created the graph from scratch, as in this case, you can add the definitions of possible edges
that the graph can store. Adding an edge will automatically adds the collections to the graph as possible sources of edges, and it will creates a collection for the edge relation

```{r step_7, echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE}
residenceGraph <- 
  residenceGraph %>% 
  define_edge("person", "lives_in", "city")

livesInCollection <- 
  sandboxArangoDb %>% 
  arango_collection("lives_in")

if(!(is.null(livesInCollection))){
  print(paste0("'Voilà, I'm an edge collection, isn't it?' ", 
               "'",livesInCollection$getType() == collection_type$EDGE,"'"))
}
```

```{r step_7_real, echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE}
residenceGraph <- 
  residenceGraph %>% 
  define_edge("person", "lives_in", "city")

livesInCollection <- 
  sandboxArangoDb %>% 
  arango_collection("lives_in")

if(!(is.null(livesInCollection))){
  print(paste0("'Voilà, I'm an edge collection, isn't it?' ", 
               "'",livesInCollection$getType() == collection_type$EDGE,"'"))
}
```

Also collections will be automatically created if not in the collection set:

```{r step_8, echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE}
residenceGraph <- 
  residenceGraph %>% 
  define_edge(cities, "had_weather", "weather")
  
weatherCollection <- 
  sandboxArangoDb %>% 
  arango_collection("weather")

if(!(is.null(weatherCollection))){
  print(paste0("'Voilà, I'm a document collection, isn't it?' ", 
               "'",weatherCollection$getType() == collection_type$DOCUMENT,"'"))
}
```

```{r step_8_real, echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE}
residenceGraph <- 
  residenceGraph %>% 
  define_edge(cities, "had_weather", "weather")
  
weatherCollection <- 
  sandboxArangoDb %>% 
  arango_collection("weather")

if(!(is.null(weatherCollection))){
  print(paste0("'Voilà, I'm a document collection, isn't it?' ", 
               "'",weatherCollection$getType() == collection_type$DOCUMENT,"'"))
}
```

But a graph isn't useful if you cannot populate with relations. Just use the aRangodb::add_edges() method and the relational operators!

```{r step_9, echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE}
residenceGraph <- 
  residenceGraph %>%
  add_edges("lives_in" %owns% edge(all.persons$john.doe %->% all.cities$London)) %>%
  add_edges("lives_in" %owns% edge(all.persons$brandon.fee %->% all.cities$Manchester, since="09/01/2016"))
```

```{r step_9_real, echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE}
residenceGraph <- 
  residenceGraph %>%
  add_edges("lives_in" %owns% edge(all.persons$john.doe %->% all.cities$London)) %>%
  add_edges("lives_in" %owns% edge(all.persons$brandon.fee %->% all.cities$Manchester, since="09/01/2016"))
```